Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota \\ {\large Senior Comps Project}

### 2010-03-01 Paper.tex
Edits to standard model and discussion sections.

    WinBUGS requires a representation of a model, data, and initial values on
    which a predetermined number of iterations of the MCMC algorithm are
    executed \cite{MacKenzie2006}.
    The data we received from the
    Minnesota Department of Natural Resources was not fixed with respect to
    our covariates and required us to implement one of the three common
    strategies used to deal with irregular data sets \cite{Spiegelhalter2003}.
    Our simulations, on the other
    hand, ran on data sets with a fixed number of observers, snow events, and
    observations after each snow event, which are much easier to implement
    within WinBUGS and R.

@@

    \subsection{Our hierarchical model}
    For our model, we specified noninformative uniform \((0, 1)\) prior
    distributions for
    parameters \(\psi\) and \(\theta\), a common, natural choice given a lack of
    formal information about the prior or expert consensus \cite{MacKenzie2006}.
    We implemented models with normal and beta prior distributions on \(p\) and
    \(E\) and limited the range of \(p\) to the interval \((0.6, 1)\) and the
    range of \(E\) to the interval \((0, 0.4)\) in both models. We set the
    burn-in time for the MCMC simulation to 10,000 iterations. Each of these
    parameters are tracked for 10,000 additional iterations and are estimated
    by the mean of the posterior distribution. The covariates of this
    model are the observer and the number of days since the last snow event.
    It was necessary to preprocess the Minnesota DNR's data to account for its
    irregularities.

    \subsection{Simulated data}
    We simulated both balanced and unbalanced data sets meant for the standard
    hierarchical model with given values of \(\psi\), \(\theta\), \(p\) for each
    observer, and \(E\) for each observer.  Balanced data sets, in this context,
    contain the same number of flights by the same observers after each snow
    event.  Naturally, unbalanced data sets do not.

    To simulate balanced data, we provided the following additional variables:
    the number of plots, the number of snow events, the number of observers, and
    the number of observations following each snow event. These variables are
    all that are required to allow us to generate data in which each snow event
    receives the same number of flights by the same number of observers, which
    greatly simplifies the accompanying hierarchical model. First, the
    simulation determines a plot's occupancy based upon the given \(\psi\)
    value.
    Then, for each day after each snow event, the simulation decides whether or
    not an otter laid down a track. For a given day after a snow event, a track
    is considered to be present if an otter lays a track down on any of the days
    after the snow event up to and including the given day. Once that has been
    decided, the given \(p\) and \(E\) values are used to assign a 0 (not
    detected) or 1 (detected).

    Because of limitations with respect to WinBUGS' syntax, simulating
    unbalanced data requires much more additional information and a reformatting
    of the data as well as the corresponding hierarchical model. While the
    resulting code's structure differs significantly from that used for balanced
    data, the underlying model remains the same.

    To determine the efficacy of the standard model over the range of possible
    \(\psi\) and \(\theta\) values, we ran simulations with several combinations
    of \(\psi\) and \(\theta\), ran the model on each simulation, and compared
    the model's estimates to the simulation's parameters (Table \ref{Simulated
    data}). The model with beta
    prior distributions on \(p\) and \(E\) performed better than the model with
    normal prior distributions on the same parameters.  Therefore, we limit
    further discussion to the beta model.

    \begin{table}
    \caption{Beta model estimates from simulated data. All simulations
    ran with 70 plots, 5 snow events, 3 observers, and for 3 days after each
    snow event. The 3 observers had \(p\) values of 0.70, 0.80, 0.90 and \(E\)
    values of 0.05, 0.20, and 0.30, respectively. Green rows represent values of
    \(\psi\) and \(\theta\) for which the model exhibits good convergence and
    the 95\% confidence interval captures the actual \(\psi\) value.  Yellow
    rows represent values for which our model exhibits either mediocre
    convergence or a bad confidence interval.  Red rows represent values for
    which our model does not converge.}
    \label{Simulated data}
    \begin{center}
    \begin{tabular}{|l|l|l|l|}
        \hline
            \(\psi\) & \(\theta\) & Convergence & 95\% CI (\(\psi\)) \\
        \hline
        \multirow{5}{*}{0.20}
            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.112, 0.305) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.142, 0.341) \\
            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.098, 0.284) \\
            & \cellcolor{Green}0.80 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.097, 0.281) \\
            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.117, 0.302) \\
        \hline
        \multirow{5}{*}{0.40}
            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.388, 0.721) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.270, 0.494) \\
            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.287, 0.512) \\
            & \cellcolor{Red}0.80 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.277, 0.507) \\
            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.279, 0.503) \\
        \hline
        \multirow{5}{*}{0.60}
            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
              \cellcolor{Yellow}(0.312, 0.581) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.400, 0.630) \\
            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.462, 0.687) \\
            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.502, 0.727) \\
            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.318, 0.549) \\
        \hline
        \multirow{5}{*}{0.80}
            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.578, 0.838) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.736, 0.915) \\
            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.587, 0.829) \\
            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.599, 0.811) \\
            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.612, 0.830) \\
        \hline
        \multirow{5}{*}{1.00}
            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
              \cellcolor{Yellow}(0.895, 0.999) \\
            & \cellcolor{Yellow}0.40 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.927, 0.999) \\
            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.949, 1.000) \\
            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Good &
              \cellcolor{Yellow}(0.802, 0.954) \\
            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.800, 0.955) \\
        \hline
    \end{tabular}
    \end{center}
    \end{table}

    Generally, the model is most effective for lower values of \(\psi\) and
    \(\theta\) but will either converge well or provide useful estimates for the
    vast majority of combinations of \(\psi\) and \(\theta\) values. Twenty-one
    of the
    model's 25 confidence intervals for \(\psi\) include its actual value.
    The model's confidence intervals for each of the
    three values of \(p\) include its actual value for all \(\psi\) and
    \(\theta\) values. On the other hand, the model accurately estimates \(E\)
    for low values of \(\theta\), but consistently overestimates \(E\) when
    \(\psi\geq0.4\) and \(\theta\geq0.6\) regardless of \(E\)'s actual value.
    This may help to explain why the model does not estimate \(\psi\) as
    successfully for higher values of \(\psi\) and \(\theta\).

@@

    \subsection{Obtaining convergence on the data}
    Once both the standard hierarchical model and the addition of a CAR prior
    distribution on \(\theta\) to the standard model exhibited convergence on
    our simulated data, we moved onto running both models on the data.
    In general, the sparseness of the data gathered in 2004 caused us to
    simplify the model in order to obtain results. Unfortunately, the
    simplification results in estimates of \(\psi\) that do not seem accurate
    and are not directly comparable to estimates of \(\psi\) on 2003's data.

    The standard model converged reliably on all data gathered in 2003, but it
    did not converge on any of the data gathered in 2004. Considering that the
    amount of data gathered in 2004 pales in comparison to the amount of data
    gathered in 2003, this result was not surprising. The vast majority of
    2004's data consists of just one observer taking observations once after
    each snow event, which causes \(p\) and \(E\) in particular to not converge,
    regardless of plot size.

    To obtain convergence on 2004's relatively sparse data, we removed \(E\),
    which adds to the model the assumption that there are no false positives
    and, consequently, simplifies it greatly. Although the model still did not
    converge on data with plot sizes of 804m and 1608m, both the Whitewater and
    Zumbro Rivers' data at the 402m plot size converged. Although none of the
    Mississippi River's 2004 data converged, we still include estimates of
    \(p\), \(\psi\), and \(\theta\) at the 402m plot size to accompany the
    estimates corresponding to the Whitewater and Zumbro Rivers. For the most
    part, however, we feel that the absence of \(E\) from the model makes the
    the estimates of \(\psi\) much less dependable.

    The addition of a CAR prior distribution on \(\theta\) to the standard model
    caused the model to fail to converge on all of the data. Although the
    standard model and the CAR model performed similarly on simulated data
    whenever both models converged, the standard model converged much more often
    than did the CAR model. This is most likely due to the fact that the CAR
    model attempts to estimate too many parameters. Whereas the standard model
    estimates between eight and twelve parameters on the actual data (\(\psi\),
    \(\theta\), the four parameters of the beta distributions corresponding to
    \(p\) and \(E\), \(p\) for each observer, and \(E\) for each observer), the
    CAR model estimates \(\theta\) for each plot. Because there are at least
    thirty-five plots in each data set (regardless of plot size), the CAR prior
    distribution adds a considerable amount of complexity to the model.

@@

    \subsection{Summary and interpretation of estimates}
    In Tables \ref{Mississippi}, \ref{Whitewater}, and \ref{Zumbro} lie our
    estimates for all of the data from
    2003 and 2004.
    With respect to the model's ability to converge on the actual data, the most
    important factors are plot size, the number of flights recorded, and the
    number of estimated parameters. In general, both plot size and the number of
    flights recorded must be sufficiently small and large, respectively, for the
    standard model with \(E\) to converge.

@@

    \subsection{Recommended sampling practices}
    To determine the relative importance of the different sampling variables
    (number of observers, number of snow events, and
    number of days since the most recent snow event), we ran the standard
    hierarchical model with false detection on simulated data with values of
    \(\psi\) and
    \(\theta\) similar to the estimates of \(\psi\) and \(\theta\) from 2003
    data. We varied the numbers of observers, snow events, and observations
    after
    each snow event. We assumed that observations after each snow event will
    occur on consecutive days, starting from the first day after the snow event.
    This information helped us to find the most efficient ways (with respect to
    number of flights) to gather aerial occupancy data given one, two, and three
    observers while still obtaining useful estimates from the model.

    The number of observers proved to be the least important of the three
    sampling variables. We were able to obtain accurate estimates regardless of
    the
    number of observers, as long as there were sufficient numbers of snow events
    and observations after each snow event. For one observer, at least three
    snow events and three observations on consecutive days after each snow event
    are required for the model to converge and give good estimates. For two or
    three observers, two events with three observations on consecutive days
    after each snow event is sufficient. Therefore, roughly nine, twelve, and
    eighteen flights are required to obtain good estimates for one, two, and
    three observers, respectively. Surprisingly, the most cost-effective method
    (as measured by the number of flights required) only requires one observer.

    The number of snow events proved to be the second-most important of the
    three sampling variables. Although two snow events are required, only with
    one
    observer does the addition of a third snow event noticeably affect the
    model's estimates.

    The number of observations after each snow event proved to be the most
    important of the three sampling variables. Because the model relies upon the
    assumption that the probability of seeing a track increases each day because
    tracks made on previous days within the same snow event can still be seen,
    this is not surprising. At least two observations after each snow event on
    consecutive days are required. Additional observations after each event help
    the model converge more quickly, but the estimates become only slightly more
    accurate.

@@

    In summary, for future data gathering trips, we strongly suggest using one
    observer, as long as they are relatively confident that they can correctly
    identify an otter track and distinguish an otter track from the tracks of
    other animals in the immediate area, since the model most accurately
    estimates \(p\) and \(E\) when their true values are high and low,
    respectively. With one observer, three snow events and three or four snow
    days' worth of flights per snow event would be sufficient. If the same
    observer flies twice (or more often) in one day, perhaps they could act as a
    substitute for an additional observer. This is because it is probable that,
    with one flight's worth of experience observing the given river, the
    observer's \(p\) and \(E\) would change due to their newfound familiarity
    with the river.

>>>>

    WinBUGS requires a representation of a model, data, and initial values on
    which a predetermined number of iterations of the MCMC algorithm are
    executed \cite{MacKenzie2006}.
    The data we received from the
    Minnesota Department of Natural Resources was unbalanced with regard to
    covariates.
    Our initial simulations, on the other
    hand, ran on data sets with a fixed number of observers, snow events, and
    observations after each snow event, which are much easier to implement
    within WinBUGS and R.

@@

    \subsection{Our hierarchical model}
    For our model, we specified noninformative uniform \((0, 1)\) prior
    distributions for
    parameters \(\psi\) and \(\theta\), a common, natural choice given a lack of
    formal information about the prior or expert consensus \cite{MacKenzie2006}.
    We implemented models with normal and beta prior distributions on \(p\) and
    \(E\) and limited the range of \(p\) to the interval \((0.6, 1)\) and the
    range of \(E\) to the interval \((0, 0.4)\) in both models. We set the
    burn-in time for the MCMC simulation to 10,000 iterations, which seemed
    necessary to handle convergence. Each of these
    parameters are tracked for 10,000 additional iterations and are estimated
    by the mean of the posterior distribution. The covariates of this
    model are the observer and the number of days since the last snow event.
    It was necessary to preprocess the Minnesota DNR's data to account for its
    irregularities.

    \subsection{Simulated data}
    We simulated both balanced and unbalanced data sets meant for the standard
    hierarchical model with given values of \(\psi\), \(\theta\), \(p\) for each
    observer, and \(E\) for each observer.  Balanced data sets, in this context,
    contain the same number of flights by the same observers after each snow
    event.

    To simulate balanced data, we provided the following additional variables:
    the number of plots, the number of snow events, the number of observers, and
    the number of observations following each snow event. These variables are
    all that are required to allow us to generate data in which each snow event
    receives the same number of flights by the same number of observers, which
    greatly simplifies the accompanying hierarchical model. For the simulation,
    we first determine each plot's occupancy status based
    based upon the given \(\psi\) value.
    Then, for each day after each snow event, the simulation decides whether or
    not an otter laid down a track. For a given day after a snow event, a track
    is considered to be present if an otter lays a track down on any of the days
    after the snow event up to and including the given day. Once that has been
    decided, the given \(p\) and \(E\) values are used to assign a 0 (not
    detected) or 1 (detected).

    Because of limitations with respect to WinBUGS' syntax, simulating
    unbalanced data requires additional information and a reformatting
    of the data as well as the corresponding hierarchical model. While the
    resulting code's structure differs from that used for balanced
    data, the underlying model remains the same.

    To determine the efficacy of the standard model over the range of possible
    \(\psi\) and \(\theta\) values, we ran simulations with several combinations
    of \(\psi\) and \(\theta\), ran the model on each simulation, and compared
    the model's estimates to the simulation's parameters (Table \ref{Simulated
    data}). The model with beta
    prior distributions on \(p\) and \(E\) performed better than the model with
    normal prior distributions on the same parameters. The beta distribution
    gives a flexible two-parameter family of distributions on \((0,1)\). We
    included two hyper-parameters (\(\alpha\) and \(\beta\)) on the beta
    distribution as part of our hierarchical modeling, which seemed to work
    well.

    \begin{table}
    \caption{Model estimates from simulated data. All simulations
    ran with 70 plots, 5 snow events, 3 observers, and for 3 days after each
    snow event. The 3 observers had \(p\) values of 0.70, 0.80, 0.90 and \(E\)
    values of 0.05, 0.20, and 0.30, respectively. Green rows represent values of
    \(\psi\) and \(\theta\) for which the model exhibits good convergence and
    the 95\% confidence interval captures the actual \(\psi\) value.  Yellow
    rows represent values for which our model exhibits either mediocre
    convergence or a bad confidence interval.  Red rows represent values for
    which our model does not converge.}
    \label{Simulated data}
    \begin{center}
    \begin{tabular}{|l|l|l|l|}
        \hline
            \(\psi\) & \(\theta\) & Convergence & 95\% CI (\(\psi\)) \\
        \hline
        \multirow{5}{*}{0.20}
            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.112, 0.305) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.142, 0.341) \\
            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.098, 0.284) \\
            & \cellcolor{Green}0.80 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.097, 0.281) \\
            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.117, 0.302) \\
        \hline
        \multirow{5}{*}{0.40}
            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.388, 0.721) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.270, 0.494) \\
            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.287, 0.512) \\
            & \cellcolor{Red}0.80 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.277, 0.507) \\
            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.279, 0.503) \\
        \hline
        \multirow{5}{*}{0.60}
            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
              \cellcolor{Yellow}(0.312, 0.581) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.400, 0.630) \\
            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.462, 0.687) \\
            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.502, 0.727) \\
            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.318, 0.549) \\
        \hline
        \multirow{5}{*}{0.80}
            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.578, 0.838) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.736, 0.915) \\
            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.587, 0.829) \\
            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.599, 0.811) \\
            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.612, 0.830) \\
        \hline
        \multirow{5}{*}{1.00}
            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
              \cellcolor{Yellow}(0.895, 0.999) \\
            & \cellcolor{Yellow}0.40 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.927, 0.999) \\
            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.949, 1.000) \\
            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Good &
              \cellcolor{Yellow}(0.802, 0.954) \\
            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.800, 0.955) \\
        \hline
    \end{tabular}
    \end{center}
    \end{table}

    Generally, the model is most effective for lower values of \(\psi\) and
    \(\theta\) but will either converge well or provide useful estimates for the
    vast majority of combinations of \(\psi\) and \(\theta\) values. Twenty-one
    of the
    model's 25 confidence intervals for \(\psi\) capture the actual value.
    The model's confidence intervals for each of the
    three values of \(p\) include its actual value for all \(\psi\) and
    \(\theta\) values. On the other hand, the model accurately estimates \(E\)
    for low values of \(\theta\), but consistently overestimates \(E\) when
    \(\psi\geq0.4\) and \(\theta\geq0.6\) regardless of \(E\)'s actual value.
    This may help to explain why the model does not estimate \(\psi\) as
    successfully for higher values of \(\psi\) and \(\theta\).

@@

    \subsection{Obtaining convergence on the data}
    Once both the standard hierarchical model and the addition of a CAR prior
    distribution on \(\theta\) to the standard model exhibited convergence on
    our simulated data, we moved on to running both models on the actual data.
    In general, the sparseness of the data gathered in 2004 caused us to
    simplify the model in order to obtain results. Unfortunately, the
    simplification results in estimates of \(\psi\) that do not seem accurate
    and are not directly comparable to estimates of \(\psi\) on 2003's data.

    The standard model converged reliably on all data gathered in 2003, but it
    did not converge on any of the data gathered in 2004. Considering that the
    amount of data gathered in 2004 pales in comparison to the amount of data
    gathered in 2003, this result was not surprising. The vast majority of
    2004 data consists of just one observer taking observations once after
    each snow event, which causes \(p\) and \(E\) in particular to not converge,
    regardless of plot size.

    To obtain convergence on 2004 data, we removed \(E\), thus disallowing false
    positives. Although the model still did not
    converge on data with plot sizes of 804m and 1608m, both the Whitewater and
    Zumbro Rivers' data at the 402m plot size converged. Although none of the
    Mississippi River's 2004 data converged, we still include estimates of
    \(p\), \(\psi\), and \(\theta\) at the 402m plot size to accompany the
    estimates corresponding to the Whitewater and Zumbro Rivers. For the most
    part, however, we feel that the absence of \(E\) from the model makes the
    the estimates of \(\psi\) less dependable.

    The addition of a CAR prior distribution on \(\theta\) to the standard model
    caused the model to fail to converge on all of the data. Although the
    standard model and the CAR model performed similarly on simulated data
    whenever both models converged, the standard model converged much more often
    than did the CAR model. This is most likely due to the fact that the CAR
    model attempts to estimate many more parameters. Whereas the standard model
    estimates between eight and twelve parameters on the actual data (\(\psi\),
    \(\theta\), the four parameters of the beta distributions corresponding to
    \(p\) and \(E\), \(p\) for each observer, and \(E\) for each observer), the
    CAR model estimates \(\theta\) for each plot. Because there are at least
    thirty-five plots in each data set (regardless of plot size), the CAR prior
    distribution adds a considerable amount of complexity to the model.

@@

    \subsection{Summary and interpretation of estimates}
    Tables \ref{Mississippi}, \ref{Whitewater}, and \ref{Zumbro} contain our
    estimates for all of the data from
    2003 and 2004.
    With respect to the model's ability to converge on the actual data, the most
    important factors are plot size, the number of flights recorded, and the
    number of estimated parameters. In general, both plot size and the number of
    flights recorded must be sufficiently small and large, respectively, for the
    standard model with \(E\) to converge.

@@

    \subsection{Recommended sampling practices}
    To determine the relative importance of the different sampling variables
    (number of observers, number of snow events, and
    number of days since the most recent snow event), we ran the standard
    hierarchical model with false detection on simulated data with values of
    \(\psi\) and
    \(\theta\) similar to the estimates of \(\psi\) and \(\theta\) from 2003
    data. We varied the numbers of observers, snow events, and observations
    after
    each snow event. We assumed that observations after each snow event will
    occur on consecutive days, starting from the first day after the snow event.
    This information helped us to find more efficient ways (with respect to
    number of flights) to gather aerial occupancy data given one, two, and three
    observers while still obtaining useful estimates from the model.

    \textbf{The number of observers appears to be the least important of the
    three
    sampling variables.} We were able to obtain accurate estimates regardless of
    the
    number of observers, as long as there were sufficient numbers of snow events
    and observations after each snow event. For one observer, at least three
    snow events and three observations on consecutive days after each snow event
    are required for the model to converge and give good estimates. For two or
    three observers, two events with three observations on consecutive days
    after each snow event is sufficient. Therefore, roughly nine, twelve, and
    eighteen flights are required to obtain good estimates for one, two, and
    three observers, respectively. Surprisingly, the most cost-effective method
    (as measured by the number of flights required) only requires one observer.

    \textbf{The number of snow events appears to be the second-most important of
    the
    three sampling variables.} Although two snow events are required, only with
    one
    observer does the addition of a third snow event noticeably affect the
    model's estimates.

    \textbf{The number of observations after each snow event appears to be the
    most
    important of the three sampling variables.} Because the model relies upon
    the
    assumption that the probability of seeing a track increases each day because
    tracks made on previous days within the same snow event can still be seen,
    this is not surprising. At least two observations after each snow event on
    consecutive days are required. Additional observations after each event help
    the model converge more quickly, but the estimates become only slightly more
    accurate.

@@

    In summary, for future data gathering trips, we suggest using one
    observer, as long as they are relatively confident that they can correctly
    identify an otter track and distinguish an otter track from the tracks of
    other animals in the immediate area, since the model most accurately
    estimates \(p\) and \(E\) when their true values are high and low,
    respectively. With one observer, three snow events and three or four snow
    days' worth of flights per snow event would be sufficient. If the same
    observer flies twice (or more often) in one day, perhaps they could act as a
    substitute for an additional observer. This is because it is probable that,
    with one flight's worth of experience observing the given river, the
    observer's \(p\) and \(E\) would change due to their newfound familiarity
    with the river.
